---
title: "<code>MotifPeeker</code><br>Benchmarking Epigenomic Profiling Methods Using Motif Enrichment"  
author: "`r rworkflows::use_badges()`" 
date: "<h4>Updated: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i></h4>"
output:
  github_document
---

```{r, echo=FALSE, include=FALSE}
pkg <- read.dcf("DESCRIPTION", fields = "Package")[1]
title <- gsub("\n"," ",read.dcf("DESCRIPTION", fields = "Title")[1])
description <- gsub("\n"," ",read.dcf("DESCRIPTION", fields = "Description")[1])
URL <- read.dcf('DESCRIPTION', fields = 'URL')[1]
owner <- strsplit(URL,"/")[[1]][4]
repo <- strsplit(URL,"/")[[1]][5]
```


# Introduction

`MotifPeeker` is used to compare and analyse datasets from epigenomic
profiling methods with motif enrichment as the key benchmark. The package
outputs an HTML report consisting of three sections:  

1. **General Metrics**: Overview of peaks-related general metrics for the
datasets (FRiP scores, peak widths and motif-summit distances). 
2. **Known Motif Enrichment Analysis**: Statistics for the frequency of
user-provided motifs enriched in the datasets. 
3. **De-Novo Motif Enrichment Analysis**: Statistics for the frequency of
de-novo discovered motifs enriched in the datasets and compared with known
motifs. 
 
<!-- If you use ``r pkg``, please cite:  -->

<!-- > `r utils::citation(pkg)$textVersion` -->

# Installation 

`MotifPeeker` uses
[`memes`](https://www.bioconductor.org/packages/release/bioc/html/memes.html)
which relies on a local install of the
[MEME suite](https://meme-suite.org/meme/). Installation instructions can be
found [here](https://www.bioconductor.org/packages/release/bioc/vignettes/
memes/inst/doc/install_guide.html).

**NOTE:** It is important that Perl dependencies associated with MEME suite are
also installed, particularly `XML::Parser`, which can be installed using the
following command in the terminal:
```
cpan install XML::Parser
```
For more information, refer to the Perl dependency section of the MEME suite
[here](https://meme-suite.org/meme/doc/install.html#prereq_perl).

Once the MEME suite and its associated Perl dependencies are installed,
`MotifPeeker` can be installed using the following code:
```R
if(!require("remotes")) install.packages("remotes")

remotes::install_github("`r paste(owner,repo,sep='/')`")
library(`r pkg`)
```
## Documentation 

### [Website](https://`r owner`.github.io/`r repo`) 
### [Getting started](https://`r owner`.github.io/`r repo`/articles/`r pkg`) 

<hr>

## Session Info

<details>
```{r}
utils::sessionInfo()
```
</details>

