---
title: "`r MotifPeeker::report_header()`"
date: "`r format(Sys.Date(), '%b-%d-%Y')`"
output: 
    html_document: 
        toc: yes
        toc_float: yes
        code_folding: hide 
        number_sections: true
        css: custom.css
params:
    peak_files:
        value: NULL
    reference_index:
        value: 1
    alignment_files:
        value: NULL
    exp_labels:
        value: NULL
    exp_type:
        value: NULL
    genome_build:
        value: NULL
    motif_files:
        value: NULL
    motif_labels:
        value: NULL
    cell_counts:
        value: NULL
    denovo_motif_discovery:
        value: NULL
    denovo_motifs:
        value: 3
    motif_db:
        value: NULL
    download_buttons:
        value: NULL
    output_dir:
        value: NULL
    use_cache:
        value: NULL
    ncpus:
        value: NULL
    debug:
        value: FALSE
    verbose:
        value: TRUE
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                    message = params$debug,
                    warning = params$debug,
                    cache = FALSE, 
                    error = params$debug,
                    results = "asis")

### Parse input ###
genome_build <- check_genome_build(params$genome_build)
result_len <- length(params$peak_files)
if (is.null(params$exp_type)) {
    exp_types <- rep("unknown", result_len)
} else {
    exp_types <- params$exp_type
}
result <- list(
    peaks = Vectorize(read_peak_file, "peak_file")(params$peak_files),
    alignments = Vectorize(Rsamtools::BamFile,
                            "file")(params$alignment_files),
    exp_labels = params$exp_labels,
    exp_type = Vectorize(format_exptype,
                        "exp_type")(exp_types)
)
user_motifs <- list(
    motifs = Vectorize(read_motif_file, "motif_file")(params$motif_files),
    motif_labels = params$motif_labels
)

```

[`MotifPeeker`](https://github.com/neurogenomics/MotifPeeker) compares different
epigenomic datasets using motif enrichment as the key metric.

# Summary {-}
## Table of Contents {-}
The report consists of the following sections:  

1. [General Metrics](#general-metrics): Overview of general metrics related to
peaks in the datasets (FRiP scores, peak widths, and motif-summit distances).

2. [Known Motif Enrichment Analysis](#motif-enrichment-analysis): Statistics on
the frequency of motifs enriched in the datasets. Also compares enriched motifs
between common and unique peaks identified in the datasets.

3. [De-Novo Motif Enrichment Analysis](#denovo-motif-analysis): Statistics on
de-novo motifs discovered between common and unique peaks identified in the
datasets. Also checks for similarity between motifs in each set and finds the
closest known motif in the [JASPAR database](https://jaspar.uio.no/downloads/)
(or the supplied database).

## Input Datasets {-}
Experimental dataset labels used:  
```{r echo = FALSE}
for (i in seq_along(result$peaks)) {
    cat(paste0("\n- **Dataset ", i, "**: ", result$exp_labels[i], "  \n"))
    cat(paste0("  - **Experiment Type**: ", result$exp_type[i], "  \n"))
    if (!is.null(params$cell_counts)) {
        cat(paste0("  - **Cell Count**: ", params$cell_counts[i], "  \n"))
    }
}
cat("\n- **Reference dataset**: ", result$exp_labels[params$reference_index], 
    paste0("(Index: ", params$reference_index, ")  \n"))
```

User-provided motifs used:  

```{r echo = FALSE}
for (i in seq_along(user_motifs$motifs)) {
    cat(paste0("- **Motif ", i, "**: ", user_motifs$motifs[[i]]@name, "  \n"))
}
```

## Command {-}
`MotifPeeker` report was generated with the following command:
```{r report_command}
cat(report_command(params))
```

## Summary Plot {-}
Lorem ipusm


# Session Info {-}
<details> 
```{r Session Info, results='markup'}
utils::sessionInfo()
```
</details> 
<hr>
